Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-09 Ben_Manuscripts/hdphmm/Draft.tex
Comments on draft.

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the infinite hidden Markov model (IHMM) to solute center-of-mass trajectories
  generated from long molecular dynamics simulations in a cross-linked inverted hexagonal
  phase lyotropic liquid crystal (LLC) membrane in order to automatically detect a
  variety of solute dynamical states which can be further explained in terms of 
  solute-membrane interactions.
% order to automatically detect and parameterize an unknown number of
%  latent vector autoregressive dynamical modes. 
  %MRS: in above sentence: try to rephrase around the physical information obtained (mechanism or properties) rather than the mathematical route to it.
  %BJC2: gave a try

  %MRS: below - it's not the solute behavior, though, it's states identified by them model. Rephrase?
  %BJC2: I think this is better?
  We group the states identified by the IHMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane. This enhancement to our understanding of the diverse range 
  of influential solute-membrane interactions allows us to hypothesize design changes 
  to LLC monomers aimed towards controlling the rates of solute passage, thus improving 
  the selective performance of LLC membranes.

@@

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms in 
  %BJC2: porous nanoscale to nanoporous? 
  %porous nanoscale 
  nanoporous materials and project complex diffusive behavior on
  long time scales.

@@

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables.
  %MRS: should explicitly point out that just calculating diffusion coefficients don't work for subdiffusive behavior, so
  %MRS need a different way to do it.
  %BJC2: added last two bullet points
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	\item In many systems, it's possible to achieve this by calculating diffusion 
  	constants.\cite{}
  	\item Unfortunately, for systems which exhibit complex subdiffusive dynamics on the
  	time scales accessible to MD simulation, one cannot calculate reliable diffusion 
  	constants since solute mean square displacements (MSDs) are non-linear.
  	\item Therefore, we are in need of a more flexible approach for modeling solute 
  	dynamics which makes less assumptions about the solute's average behavior.
  	%BJC2: I'm moving these towards the end
%  	\item There are two main challenges associated with this goal.
%  	%since complex and
%  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
%  	%us from easily relating chemically dependent solute motion to experimental time scales.
%  	\item First, we need a simple way to identify and parameterize the fluctuations
%  	characteristic to dominant solute-membrane interactions.
%  	\item Second, we need a way to propagate this behavior far beyond the time scales
%  	afforded by MD.
  \end{itemize}

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.
    %MRS: Then say it hasn't been used in this context (do we know of any, if not say so).
    \item The IHMM has not been used much for MD trajectory analysis.

@@

  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to the dynamical behavior which give rise to those types
  of observations. The probability of transitioning to a state based on the current
  state is mathematically defined in terms of an $n\times n$ transition probability
  matrix, $T$, where $n$ is the number of states. Unfortunately, standard HMMs 
  require $n$ to be known \textit{a priori}. One can partially overcome this by 
  testing a range of numbers of hidden states and determining which is the best 
  representation of the data.~\cite{pohle_selecting_2017}

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.~\cite{fox_bayesian_2010} Using some 
  base probability distribution, $H$, a Dirichlet process (DP) generates discrete 
  distributions, $G_0$, over a countably infinite number of probability measures:
  \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
  \end{equation}
  where the $\theta_k$ are values drawn from the base distribution and the weights
  $\beta_k$ come from a stick-breaking process parameterized by the concentration 
  parameter $\gamma$ (equivalently referred to as GEM($\gamma$)).~\cite{halmos_random_1944} In shorthand, this
  specification can be expressed as $G_0 \sim DP(\gamma, H)$. The concentration 
  parameter, $\gamma$, expresses one's confidence in the base distribution $H$.
  We use a uniform base distribution for $H$. When $\gamma\to 0$, the first 
  weight of $G_0$, $\beta_1$, approaches unity and for $\gamma\to\infty$, the weights
  become uniform and $G_0$ closely resembles $H$. Each row, $G_j$, of the transition 
  matrix is produced by drawing from a DP specified using the $\beta$ vector as a 
  discrete base distribution and a separate concentration parameter, $\alpha$.
  \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
  \end{equation}
  This hierarchical specification ensures that the transition probabilities in 
  each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
  Once the model has converged only a finite number of states will have significant
  sampling.

@@

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  % each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of Equation~\ref{eqn:var}. 
  We detected equilibration 
  of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

@@

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) due to the system's radial 
  symmetry. We converted the $x$ and $y$ center-of-mass coordinates to $r$. 
  We fixed the state sequence from the 3D parameterization and applied the 
  inference component of the IHMM procedure to the cylindrical trajectories 
  in order to estimate the VAR(1) parameters in terms of $r$ and $z$. Since
  we fixed the state sequence, and we are using conjugate priors, the VAR 
  parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
  Therefore, we only ran the inference procedure for 100 iterations. 
  % MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
  % MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)

  We defined the final parameters of each state as the mean of the parameters 
  from each iteration recorded after the equilibration time point. We found the 
  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
  dimensions as the equilibration time point.

@@

  %MRS: maybe add subsubsections naming/labeling each part of the process so that people have a sense of the steps,
  %MRS: makes it easier to read.
  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size. For each solute studied, we identified 200-325
  independent states, each with separate VAR(1) parameters. Many of these states
  exhibit very similar dynamical behavior except their mean levels in $r$ and $z$
  are different.

@@

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the radial means, $\mu_r$, 
  and the self-transition probabilities, $T_{ii}$, of each state for a total of 6 clustering
  dimensions. The means in $r$ are likely meaningful since the membrane is inhomogeneous, 
  transitioning from hydrophilic pores to hydrophobic tails. To cluster on $T_{ii}$, we first
  cast it in terms of the expected dwell times. The expected value of the number of sequential
  self-transitions is simply $\frac{1}{1 - T_{ii}}$. This relationship correctly implies that 
  dwell times approach infinity as $T_{ii}$ approaches 1. There we found the most success with
  agglomerative clustering by linearizing this relationship and clustering on $-\log(1 - T_{ii})$
  (see Section\ref{S-section:selfT_cluster} of the Supporting Information for a graphical
  illustration).

@@

  %We developed a way to 
  We measure the local density of the membrane at arbitrary
  points in the unit cell as follows. We histogrammed the positions in three dimensions
  Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the +/- x, y and z directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011} We then used a regular
  grid interpolator in order to allow interpolation at arbitrary points within the grid.~\cite{virtanen_scipy_2020}

  We quantified solute motion using the time averaged mean squared displacement (MSD)
  of the solute center of mass trajectories.~\cite{meroz_toolbox_2015} The time-averaged MSD measures all observed
  displacements over time lag $\tau$:
  \begin{equation}
  	\overline{z^2(\tau)} = \dfrac{1}{T - \tau}\int_{0}^{T - \tau} (z(t + \tau) - z(t))^2 dt
  \label{eqn:tamsd}
  \end{equation}
  where T is the length of the trajectory. We reported 1$\sigma$ confidence intervals 
  based on the results of 200 bootstrap trials. For each trail, we calculated the mean of 
  $n$ random trajectories chosen with replacement from the pool of trajectories, where 
  $n$ is the number of trajectories.

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{The MD MSDs (black) of methanol and urea are under-predicted by 
  the IHMM (orange), while the predictions match within uncertainty for ethylene
  glycol and urea.}\label{fig:unclustered_msds}
  \end{figure}

  \noindent The IHMM under-predicts MD MSD due to its tendency to favor fewer states.\cite{}
  \begin{itemize}
    \item Disjoint segments with similar VAR parameters, but different lengths of 
    entrapment can be parameterized as the same state.
    \item In Figure~\ref{fig:dwell_distributions}, we compare the overall dwell 
    time distributions generated directly from MD simulations to that generated from 
    realizations of the IHMM. 
    \item Some of the probability density of short trapping times in the MD simulations
    gets shifted to medium length trapping times in realizations of the IHMM.
    \item In Figure~\ref{fig:dwell_distributions_modT} and~\ref{fig:msd_modT}, we show 
    that by modifying the transition matrix to encourage transitions towards states with
    shorter dwell times, we can more closely reproduce the dwell time distribution and 
    MSD of MD.
    \item This is because most solute motion is a consequence of transitions between 
    trapped states.
  \end{itemize}

@@

  Also apparent from Figure~\ref{fig:unclustered_msd_MET_shortlag} is that the 
  curvature of the MD MSD is far more gradual than that generated by the IHMM.
  \begin{itemize}
    \item This is because the VAR(1) model is only correlated to its previous
    fluctuation while the MD simulations suggest that correlation persists for
    at least tens of nanoseconds.
    % BJC2: could probably use a correlation function to suggest the AR order and stick the plot in SI
    \item One may be able to reproduce the curvature at short time lags with
    higher order VAR models, but it becomes a much more difficult parameterization
    that would require significantly more data.
    \item Since we are interested in longer time scale behavior, it is not an 
    issue of great concern.
  \end{itemize}

@@

  %BJC2: the exact figure will change, but most of what I say shouldn't
  \begin{figure}
  \centering
  \begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\textwidth]{mechanism_map.png}
  \caption{}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{monomer_oxygens.png}
  \caption{}
  \end{subfigure}
  %MRS: a lot of this caption test is duplicated from the main text. I would suggest reducing the duplication.
  %BJC: you think better to have more detail in text or in caption?
  \caption{(a) We can learn a significant amount of detail about solute motion by viewing 
  solute trajectories color-coded according to their clustered dynamical behavior alongside
  plots of the dominant molecular interactions and descriptions of the membrane's local 
  environment. In the plot above, we show the radial and axial coordinates of a single
  methanol molecule over the course of the equilibrated portion of its MD simulation.
  Disjoint segments of the same color imply the solute behaves similarly at different 
  points of the trajectory because their VAR parameters place them in the same cluster. 
  There are eight distinct dynamical modes exhibited by this trajectory. In order
  of their first appearance, they are colored dark yellow, red, light blue, light green,
  orange, blue, teal and lime green. On the top of the plot are bars which represent different
  physical interactions and the membrane's local environment as a function of time. The top 
  bar is colored green at time points where solutes are associated with sodium. Methanol 
  associates with sodium ions relatively infrequently. The next bar down is colored when hydrogen
  bonds exist. The colors match the enlarged oxygen atoms highlighted in (b). Blue slivers
  correspond to hydrogen bonds donated to the monomer's carboxylate head group. Green slivers
  correspond to  hydrogen bonds donated to the ether linkages between the head groups
  and tails. Red slivers correspond to hydrogen bonds donated to the oxygens on the ends
  of the monomer tails. Methanol exhibits significant hydrogen bonding with all parts of
  the monomer. Which part of the monomer methanol hydrogen bonds to is heavily correlated
  to the solute's $r$ coordinate. Finally, the bar with the blue gradient represents the local 
  membrane density based on the solute's ($r$, $z$) coordinate. Darker shades of blue
  correspond to higher membrane densities. The density appears to be higher when methanol
  motion is restricted (e.g. from 75 to 550 ns). It is relatively low in areas where methanol's
  position fluctuates significantly (e.g. from 550 to 700 ns). 
  }\label{fig:mechanism_map}
  \end{figure}

@@

  It is clear that there is a wealth of information in these plots, but they don't give
  the complete picture needed in order to formulate LLC membrane design principles. One
  can analyze each segment sequentially and form a clear picture of this instance of 
  methanol's behavior without the need to visualize the trajectory with software like VMD.~\cite{humphrey_vmd:_1996}
  However, it is still possible to gain a more general understanding of methanol's average
  behavior.

@@

  The size and chemical functionality of the solutes dictates the interactions which
  influence solute transport. In Figure~\ref{fig:hbonds_assoc_summary}, we show that
  the solutes donate hydrogen bonds and associate with sodium ions to varying degrees.
  We also show the solute MSDs after a 1000 ns time lag. It is not immediately obvious
  why the MSDs follow this trend. The states identified by the IHMM shed
  some light on the reasoning.

>>>>

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the infinite hidden Markov model (IHMM) to solute center-of-mass trajectories
  generated from long molecular dynamics simulations in a cross-linked inverted hexagonal
  phase lyotropic liquid crystal (LLC) membrane in order to automatically detect a
  variety of solute dynamical states which can be further explained in terms of 
  solute-membrane interactions.

  We group the states identified by the IHMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane.

@@

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms in  nanoporous materials and project complex diffusive behavior on
  long time scales.  This enhancement to our understanding of the diverse range 
  of solute behavior allows us to hypothesize design changes 
  to LLC monomers aimed towards controlling the rates of solute passage, thus improving 
  the selective performance of LLC membranes.

@@

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables.
  %MRS: should explicitly point out that just calculating diffusion coefficients don't work for subdiffusive behavior, so
  %MRS need a different way to do it.
  %BJC2: added last two bullet points
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	\item In many systems, it's possible to achieve this by calculating diffusion 
  	constants.\cite{}
  	\item Unfortunately, for systems which exhibit complex subdiffusive dynamics on the
  	time scales accessible to MD simulation, one cannot calculate reliable diffusion 
  	constants since solute mean square displacements (MSDs) are non-linear 
        %MRS2: added
        on timescales accessible to molecular simulation.
  	\item Therefore, we are in need of a more flexible approach for modeling solute 
  	dynamics which makes less assumptions about the solute's average behavior.
  	%BJC2: I'm moving these towards the end
%  	\item There are two main challenges associated with this goal.
%  	%since complex and
%  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
%  	%us from easily relating chemically dependent solute motion to experimental time scales.
%  	\item First, we need a simple way to identify and parameterize the fluctuations
%  	characteristic to dominant solute-membrane interactions.
%  	\item Second, we need a way to propagate this behavior far beyond the time scales
%  	afforded by MD.
  \end{itemize}

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.
    %MRS: Then say it hasn't been used in this context (do we know of any, then cite, if none we have found, say so.)
    \item The IHMM has not been used much for MD trajectory analysis.

@@

  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to 
  % MRS2: added - ``states'' aren't quite ``behavior''
  segments of the trajectory with similar
  dynamical behavior which gives rise to those types
  of observations. The probability of transitioning to a state based on the current
  state is 
  %MRS2: redundant
  % mathematically 
  defined in terms of an $n\times n$ transition probability
  matrix, $T$, where $n$ is the number of states. Unfortunately, standard HMMs 
  require $n$ to be known \textit{a priori}. One can partially overcome this by 
  testing a range of numbers of hidden states and determining which is the best 
  representation of the data.~\cite{pohle_selecting_2017}

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  %MRS2: this is fine.
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.~\cite{fox_bayesian_2010} Using some 
  base probability distribution, $H$, a Dirichlet process (DP) generates discrete 
  distributions, $G_0$, over a countably infinite number of probability measures:
  \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
  \end{equation}
  where the $\theta_k$ are values drawn from the base distribution and the weights
  $\beta_k$ come from a stick-breaking process parameterized by the concentration 
  parameter $\gamma$ (equivalently referred to as GEM($\gamma$)).~\cite{halmos_random_1944} In shorthand, this
  specification can be expressed as $G_0 \sim DP(\gamma, H)$. The concentration 
  parameter, $\gamma$, expresses one's confidence in the base distribution $H$.
  We use a uniform base distribution for $H$. When $\gamma\to 0$, the first 
  weight of $G_0$, $\beta_1$, approaches unity and for $\gamma\to\infty$, the weights
  become uniform and $G_0$ closely resembles $H$. Each row, $G_j$, of the transition 
  matrix is produced by drawing from a DP specified using the $\beta$ vector as a 
  discrete base distribution and a separate concentration parameter, $\alpha$.
  \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
  \end{equation}
  This hierarchical specification ensures that the transition probabilities in 
  each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
  Once the model has converged only a finite number of states will have significant
  sampling.

@@

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  % each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of Equation~\ref{eqn:var}. 
  We detected equilibration 
  of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

@@

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) due to the system's radial 
  symmetry. We converted the $x$ and $y$ center-of-mass coordinates to $r$. 
  We fixed the state sequence from the 3D parameterization and applied the 
  inference component of the IHMM procedure to the cylindrical trajectories 
  in order to estimate the VAR(1) parameters in terms of $r$ and $z$. Since
  we fixed the state sequence, and we are using conjugate priors, the VAR 
  parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
  Therefore, we only ran the inference procedure for 100 iterations. 
  % MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
  % MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)
  % MRS2: As I mentioned before, the issue is if the likelihood for the covariance matrix is properly defined for radial diffusion.  We can check over the equations.  It likely wouldn't change the qualitative results much, but should ne checked.  Would not affect the clustering.
  We defined the final parameters of each state as the mean of the parameters 
  from each iteration recorded after the equilibration time point. We found the 
  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
  dimensions as the equilibration time point.

@@

  %MRS2: maybe add subsubsections naming/labeling each part of the process so that people have a sense of the steps,
  %MRS2: makes it easier to read.
  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size. For each solute studied, we identified 200-325
  independent states, each with separate VAR(1) parameters. Many of these states
  exhibit very similar dynamical behavior except their mean levels in $r$ and $z$
  are different.

@@

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the radial means, $\mu_r$, 
  and the self-transition probabilities, $T_{ii}$, of each state for a total of 6 clustering
  dimensions. The means in $r$ are likely meaningful since the membrane is 
  %MRS2: added
  radially inhomogeneous from each pore, 
  transitioning from hydrophilic pores to hydrophobic tails. 
  %MRS2: 
  However, the channels are isotropic in $z$ so we do not cluster on $\mu_z$.
  To cluster on $T_{ii}$, we first
  cast it in terms of the expected dwell times. The expected value of the number of sequential
  self-transitions is simply $\frac{1}{1 - T_{ii}}$. This relationship correctly implies that 
  dwell times approach infinity as $T_{ii}$ approaches 1. There we found the most success with
  agglomerative clustering by linearizing this relationship and clustering on $-\log(1 - T_{ii})$
  (see Section\ref{S-section:selfT_cluster} of the Supporting Information for a graphical
  illustration).

@@

  %We developed a way to 
  We measure the local density of the membrane at arbitrary
  points in the unit cell as follows. We histogrammed the positions in three dimensions
  Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the $\pm$$x$, $y$ and $z$ directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011} We then used a regular
  grid interpolator in order to allow interpolation at arbitrary points within the grid.~\cite{virtanen_scipy_2020}

  We quantified solute motion using the time averaged mean squared displacement (MSD)
  of the solute center of mass trajectories.~\cite{meroz_toolbox_2015} The time-averaged MSD measures all observed
  displacements over time lag $\tau$:
  \begin{equation}
  	\overline{z^2(\tau)} = \dfrac{1}{T - \tau}\int_{0}^{T - \tau} (z(t + \tau) - z(t))^2 dt
  \label{eqn:tamsd}
  \end{equation}
  where T is the length of the trajectory. We reported $1 \sigma$ confidence intervals 
  based on the results of 200 bootstrap trials. For each trail, we calculated the mean of 
  $n$ random trajectories chosen with replacement from the pool of trajectories, where 
  $n$ is the number of trajectories.

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{The MD MSDs (black) of methanol and urea are under-predicted by 
  the IHMM (orange), while the predictions match within uncertainty for ethylene
  glycol and urea.}\label{fig:unclustered_msds}
  %MRS2: to save space, could consider the graphs in 1x4 across the top, with a single legend (since all have the same), and slightly thicker lines  %There are only 2 lines per plot, so it's not too noisy to be that size
  \end{figure}

  %MRS2: I'd make this a separate subsection ``Deviations of the IMMM from molecular motion'' or something like that.
  \noindent The IHMM under-predicts MD MSD due to its tendency to favor fewer states.\cite{}
  \begin{itemize}
    \item Disjoint segments with similar VAR parameters, but different lengths of 
    entrapment can be parameterized as the same state.
    \item In Figure~\ref{fig:dwell_distributions}, we compare the overall dwell 
    time distributions generated directly from MD simulations to that generated from 
    realizations of the IHMM. 
    \item Some of the probability density of short trapping times in the MD simulations
    gets shifted to medium length trapping times in realizations of the IHMM.
    \item In Figure~\ref{fig:dwell_distributions_modT} and~\ref{fig:msd_modT}, we show 
    that by modifying the transition matrix to encourage transitions towards states with
    shorter dwell times, we can more closely reproduce the dwell time distribution and 
    MSD of MD.
    \item This is because most solute motion is a consequence of transitions between 
    trapped states.
  \end{itemize}

@@

  %MRS2: reference not working?
  Also apparent from Figure~\ref{fig:unclustered_msd_MET_shortlag} is that the 
  curvature of the MD MSD is far more gradual than that generated by the IHMM.
  \begin{itemize}
    \item This is because the VAR(1) model is only correlated to its previous
    fluctuation while the MD simulations suggest that correlation persists for
    at least tens of nanoseconds.
    % BJC2: could probably use a correlation function to suggest the AR order and stick the plot in SI
    \item One may be able to reproduce the curvature at short time lags with
    higher order VAR models, but it becomes a much more difficult parameterization
    that would require significantly more data.
    \item Since we are interested in longer time scale behavior, it is not an 
    issue of great concern.
  \end{itemize}

@@

  %BJC2: the exact figure will change, but most of what I say shouldn't
  \begin{figure}
  \centering
  \begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\textwidth]{mechanism_map.png}
  \caption{}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{monomer_oxygens.png}
  \caption{}
  \end{subfigure}
  %MRS: a lot of this caption test is duplicated from the main text. I would suggest reducing the duplication.
  %BJC: you think better to have more detail in text or in caption?
  %MRS2: at a certain point, it should be in the text.  In the caption, I would just state what the point of the plot is, 
  % and define the colors/data/etc.  But full explanation would be better in the text. Captions should be scannable quickly,
  % based on how people process papers, and this should be significantly shorter for readability. 
 \caption{(a) We can learn a significant amount of detail about solute motion by viewing 
  solute trajectories color-coded according to their clustered dynamical behavior alongside
  plots of the dominant molecular interactions and descriptions of the membrane's local 
  environment. In the plot above, we show the radial and axial coordinates of a single
  methanol molecule over the course of the equilibrated portion of its MD simulation.
  Disjoint segments of the same color imply the solute behaves similarly at different 
  points of the trajectory because their VAR parameters place them in the same cluster. 
  There are eight distinct dynamical modes exhibited by this trajectory. In order
  of their first appearance, they are colored dark yellow, red, light blue, light green,
  orange, blue, teal and lime green. On the top of the plot are bars which represent different
  physical interactions and the membrane's local environment as a function of time. The top 
  bar is colored green at time points where solutes are associated with sodium. Methanol 
  associates with sodium ions relatively infrequently. The next bar down is colored when hydrogen
  bonds exist. The colors match the enlarged oxygen atoms highlighted in (b). Blue slivers
  correspond to hydrogen bonds donated to the monomer's carboxylate head group. Green slivers
  correspond to  hydrogen bonds donated to the ether linkages between the head groups
  and tails. Red slivers correspond to hydrogen bonds donated to the oxygens on the ends
  of the monomer tails. Methanol exhibits significant hydrogen bonding with all parts of
  the monomer. Which part of the monomer methanol hydrogen bonds to is heavily correlated
  to the solute's $r$ coordinate. Finally, the bar with the blue gradient represents the local 
  membrane density based on the solute's ($r$, $z$) coordinate. Darker shades of blue
  correspond to higher membrane densities. The density appears to be higher when methanol
  motion is restricted (e.g. from 75 to 550 ns). It is relatively low in areas where methanol's
  position fluctuates significantly (e.g. from 550 to 700 ns). 
  }\label{fig:mechanism_map}
  \end{figure}

  %MRS2: suggest subsubcaptions/italics at the start of parentheses to make it easier for a reader to see the structure.
  %MRS2: like \textit{hydrogen bonding patterns}: etc.

@@

  %MRS2: not entirely clear about the purpose of this paragraph? Seems to be just saying it's possible to learn more?
  %MRS2: should clarify.
  It is clear that there is a wealth of information in these plots, but they don't give
  the complete picture needed in order to formulate LLC membrane design principles. One
  can analyze each segment sequentially and form a clear picture of this instance of 
  methanol's behavior without the need to visualize the trajectory with software like VMD.~\cite{humphrey_vmd:_1996}
  However, it is still possible to gain a more general understanding of methanol's average
  behavior.

@@

  The size and chemical functionality of the solutes dictates the interactions which
  influence solute transport. In Figure~\ref{fig:hbonds_assoc_summary}, we show that
  the solutes donate hydrogen bonds and associate with sodium ions to varying degrees.
  We also show the solute MSDs after a 1000 ns time lag. It is not immediately obvious
  why the MSDs follow this trend. 
  %MRS2: be more precise what ``this trend'' is.
  The states identified by the IHMM shed
  some light on the reasoning.
